---
output:
  html_document:
    self_contained: true
    css: _shared.css
    theme: null
    highlight: null
    mathjax: null
    pandoc_args: ["--metadata=title:Correlations — Find out more"]
params:
  cache_dir: "www/docs/_cache"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE,
                      dev = "png", dpi = 110)
runs <- readRDS(file.path(params$cache_dir, "correlations_runs.rds"))

.compare_country <- function(cn) {
  modes <- c("none", "preset", "timeseries", "manual")
  out <- do.call(rbind, lapply(modes, function(md) {
    r <- runs[[paste(cn, md, sep = "_")]]$summary
    co2e <- r[r$variable == "total_co2e", ]
    data.frame(
      Mode      = c(none = "No correlations",
                    preset = "Structural defaults",
                    timeseries = "Time-series (auto)",
                    manual = "Manual (example template)")[md],
      Mean_t    = round(co2e$mean, 0),
      CI_low_t  = round(co2e$lo, 0),
      CI_high_t = round(co2e$hi, 0),
      MoE_pct   = round(co2e$moe_pct, 1),
      stringsAsFactors = FALSE
    )
  }))
  rownames(out) <- NULL
  out
}
.fmt_t <- function(x) format(round(x), big.mark = ",")

# Iman–Conover for small inline scatter plots. Pure R, no app dependencies,
# fast on n = 1000.
.ic_pair <- function(n, rho, marg1 = rnorm, marg2 = rnorm) {
  x1 <- marg1(n); x2 <- marg2(n)
  z  <- MASS::mvrnorm(n, mu = c(0, 0),
                       Sigma = matrix(c(1, rho, rho, 1), 2, 2))
  # rank-match each marginal to the ranks in z
  data.frame(x = x1[rank(z[, 1])], y = x2[rank(z[, 2])])
}
```

We talk a lot about correlations in this tool. This page is a slow walk through what they actually do, with pictures.

The whole idea fits in one sentence: when two of your input parameters tend to move together, your Monte Carlo should know about it. If you let them move independently when they shouldn't, you sample combinations that don't exist in nature, and the result you get back is a little bit fake.

## What "correlation" looks like

Pick two parameters. Draw 1000 samples of each. Now plot them as a cloud — one dot per Monte Carlo iteration, the first parameter on the x-axis, the second on the y-axis. Here is what changes when you turn on a correlation between them:

```{r fig-scatter-normals, fig.width = 9, fig.height = 3.2, fig.alt = "Three scatter plots showing independent, positively, and negatively correlated bivariate normal samples"}
set.seed(7)
par(mfrow = c(1, 3), mar = c(3.2, 3.2, 2.6, 0.6), mgp = c(2, 0.7, 0))
for (rho in c(0, 0.7, -0.7)) {
  z <- MASS::mvrnorm(1000, mu = c(0, 0),
                      Sigma = matrix(c(1, rho, rho, 1), 2, 2))
  ttl <- if (rho == 0) "ρ = 0    (independent)"
         else if (rho > 0) sprintf("ρ = +%.1f  (move together)", rho)
         else sprintf("ρ = %.1f  (move opposite)", rho)
  plot(z, pch = 16, col = adjustcolor("#2D6A4F", 0.4),
       cex = 0.5, xlab = "parameter 1", ylab = "parameter 2",
       main = ttl, cex.main = 0.95, cex.lab = 0.85, cex.axis = 0.8,
       xlim = c(-3.5, 3.5), ylim = c(-3.5, 3.5), asp = 1)
  abline(h = 0, v = 0, col = "grey85")
}
```

Look at the middle picture. When parameter 1 is on the right (high), parameter 2 also tends to be on top (high). Same on the left: low and low. The cloud is tilted upward.

In the right-hand picture, it's the opposite: high parameter 1 means low parameter 2. The cloud tilts downward.

Notice what isn't changing. Each individual parameter's distribution is identical across all three pictures — same range on the x-axis, same range on the y-axis. Correlation doesn't change either parameter on its own. It only changes which combinations are likely.

## The trick the tool uses

In real inventory parameters, your distributions aren't usually nice symmetric normals. Body weight is roughly symmetric. Milk yield has a long right tail. Methane conversion factor Y~m~ is bounded between 5 and 8%. You can't just use the standard correlated-normals recipe and call it a day.

The tool uses a method called Iman–Conover. It samples each parameter from its own distribution (PERT, normal, lognormal, beta — whatever you specified for that row in the Parameters sheet), then re-orders the samples so that the *ranks* of one parameter match the ranks of the other to the desired correlation. The marginal distributions stay exactly what you asked for; only the joint pattern changes.

Here's a side-by-side. Two parameters with the same skewed Beta distribution. On the left: independent. On the right: ρ = −0.5 via Iman–Conover.

```{r fig-iman-conover, fig.width = 8, fig.height = 4.2, fig.alt = "Two scatter plots showing two Beta-distributed parameters under independence and under negative rank correlation"}
set.seed(13)
n <- 1500
marg <- function(n) qbeta(runif(n), 2, 5)  # right-skewed
ind  <- data.frame(x = marg(n), y = marg(n))
ic   <- .ic_pair(n, rho = -0.5, marg, marg)
par(mfrow = c(1, 2), mar = c(3.5, 3.5, 2.6, 0.6), mgp = c(2, 0.7, 0))
plot(ind, pch = 16, col = adjustcolor("#2D6A4F", 0.4), cex = 0.5,
     xlab = "parameter 1", ylab = "parameter 2",
     main = "Independent (ρ = 0)", cex.main = 0.95, cex.lab = 0.9,
     xlim = c(0, 1), ylim = c(0, 1), asp = 1)
plot(ic,  pch = 16, col = adjustcolor("#C0392B", 0.4), cex = 0.5,
     xlab = "parameter 1", ylab = "parameter 2",
     main = "Iman–Conover, ρ = −0.5", cex.main = 0.95, cex.lab = 0.9,
     xlim = c(0, 1), ylim = c(0, 1), asp = 1)
```

The right-hand cloud tilts downward, but both pictures have the same lopsided shape on each axis. That's the whole point: the *shape* of each parameter is preserved, only the *joint pattern* changes.

## Why the IPCC equations care

Now bring this back to cattle methane. The IPCC enteric-CH~4~ equation multiplies several inputs together, including digestible energy `DE` and the methane conversion factor `Y_m`. These two are biologically anti-correlated: when feed quality goes up (higher `DE`, more energy absorbed), less methane is wasted (lower `Y_m`). When feed quality goes down, the rumen ferments more and `Y_m` climbs.

If you let the Monte Carlo sample these two independently, half your iterations land on combinations that don't exist — a cow on excellent feed (high `DE`) somehow also producing massive methane (high `Y_m`). The product `DE × Y_m` for those impossible cows is unrealistically high.

What you should be doing: tie `DE` and `Y_m` together at ρ = −0.5, and the impossible combinations vanish. Watch what happens to the distribution of the product:

```{r fig-de-ym-product, fig.width = 8, fig.height = 3.6, fig.alt = "Two histograms showing the DE x Ym product under independence and under negative correlation"}
set.seed(31)
n <- 5000
# PERT-ish samples around Country X central values
de_marg <- function(n) mc2d::rpert(n, min = 40, mode = 55, max = 70, shape = 4)
ym_marg <- function(n) mc2d::rpert(n, min = 5.5, mode = 6.5, max = 7.5, shape = 4)
ind <- data.frame(de = de_marg(n), ym = ym_marg(n))
cor <- .ic_pair(n, rho = -0.5, de_marg, ym_marg)
names(cor) <- c("de", "ym")
prod_ind <- ind$de * ind$ym
prod_cor <- cor$de * cor$ym
xlim <- range(c(prod_ind, prod_cor))
par(mfrow = c(1, 2), mar = c(3.7, 3.7, 2.5, 0.6), mgp = c(2.2, 0.7, 0))
hist(prod_ind, breaks = 50, col = "#94A3B8", border = "white",
     freq = FALSE, xlim = xlim, ylim = c(0, 0.012),
     main = "DE × Yₘ — independent",
     xlab = "DE × Yₘ", ylab = "density",
     cex.main = 0.95, cex.lab = 0.9)
abline(v = mean(prod_ind), col = "black", lwd = 2, lty = 2)
hist(prod_cor, breaks = 50, col = "#2D6A4F", border = "white",
     freq = FALSE, xlim = xlim, ylim = c(0, 0.012),
     main = "DE × Yₘ — correlated ρ = −0.5",
     xlab = "DE × Yₘ", ylab = "density",
     cex.main = 0.95, cex.lab = 0.9)
abline(v = mean(prod_cor), col = "black", lwd = 2, lty = 2)
```

Two things to read off the picture:

The means (dashed lines) are almost identical. Correlations don't shift your central estimate. They never have. They never will. Anyone who tells you correlations change the headline emission number is mistaken.

The right-hand histogram is *narrower*. Negative correlation pulls the extreme combinations toward each other and shrinks the spread of the product. If you'd been using independent sampling for this pair, you would have overstated your uncertainty. Positive correlations do the opposite: they widen the spread of products.

Across the full IPCC equation chain, with seven correlated pairs and a dozen other independent inputs, these effects mostly partially-cancel, and the net result is usually a small widening of your final 95% MoE — a few percentage points. The mean of the final emission total stays put.

## Country X (smallholder dairy)

Country X is the built-in dairy example: 500,000 head, milking cows on improved pasture and concentrate supplement, `DE ≈ 55%`, `Milk ≈ 4 kg/lactating cow/day`, `Y_m ≈ 6.5%`. It's intensive enough that diet quality and methane conversion both vary substantially — exactly the system where the `DE ↔ Y_m` anti-correlation has the most room to bite.

Each row below is a full 3000-iteration simulation of Country X. The only thing that changes is which correlation mode is selected.

```{r tbl-x}
tx <- .compare_country("country_x")
knitr::kable(
  data.frame(
    Mode = tx$Mode,
    `Mean total CO₂eq (t)` = .fmt_t(tx$Mean_t),
    `95% CI lower (t)`     = .fmt_t(tx$CI_low_t),
    `95% CI upper (t)`     = .fmt_t(tx$CI_high_t),
    `95% MoE (%)`          = sprintf("%.1f", tx$MoE_pct),
    check.names = FALSE
  ),
  align = c("l", "r", "r", "r", "r")
)
```

The mean is the same to within Monte Carlo noise. The 95% MoE shifts by a couple of percentage points when you turn correlations on. Most of that shift comes from `DE ↔ Y_m` — the strongest anti-correlation in the preset. The other six preset pairs contribute smaller, mostly partially-cancelling effects.

If you tick the "Compare with/without correlations" checkbox on Tab 5 before running, you'll get a small bar chart on the Results tab showing the green-versus-grey MoEs side by side. That's the cleanest way to see the effect on *your* inventory, not just on Country X.

## Country Y (extensive pastoral, no dairy)

Country Y is 2.4 million head of beef cattle on rangeland. `DE` is lower (around 50%, with a wider uncertainty bound), milk yield is essentially zero (calves only), and the population is dominated by adult animals on year-round grazing.

```{r tbl-y}
ty <- .compare_country("country_y")
knitr::kable(
  data.frame(
    Mode = ty$Mode,
    `Mean total CO₂eq (t)` = .fmt_t(ty$Mean_t),
    `95% CI lower (t)`     = .fmt_t(ty$CI_low_t),
    `95% CI upper (t)`     = .fmt_t(ty$CI_high_t),
    `95% MoE (%)`          = sprintf("%.1f", ty$MoE_pct),
    check.names = FALSE
  ),
  align = c("l", "r", "r", "r", "r")
)
```

The same `DE ↔ Y_m` correlation is doing the same thing here, but Country Y's MoEs are wider in absolute terms across all four modes — because each individual parameter has wider uncertainty bounds in a pastoral system that's less well surveyed. The relative effect of correlations (the few-percentage-point shift you see going down the column) is similar to Country X. The biology is the same; the per-parameter uncertainty is what differs.

Here's the headline shift side by side:

```{r fig-country-shift, fig.width = 7, fig.height = 3.5, fig.alt = "Bar chart comparing 95% MoE under no correlations vs structural-defaults preset, for Country X and Country Y"}
x_none   <- runs$country_x_none$summary
x_preset <- runs$country_x_preset$summary
y_none   <- runs$country_y_none$summary
y_preset <- runs$country_y_preset$summary
get_moe <- function(s) s[s$variable == "total_co2e", "moe_pct"]
df <- data.frame(
  country = rep(c("Country X (dairy)", "Country Y (pastoral)"), each = 2),
  mode    = rep(c("No correlations", "Structural defaults"), 2),
  moe     = c(get_moe(x_none), get_moe(x_preset),
              get_moe(y_none), get_moe(y_preset))
)
par(mar = c(2.6, 4, 2.2, 0.6), mgp = c(2.6, 0.7, 0))
bp <- barplot(matrix(df$moe, ncol = 2), beside = TRUE,
              names.arg = c("Country X (dairy)", "Country Y (pastoral)"),
              col = c("#94A3B8", "#2D6A4F"),
              border = "white", ylab = "95% MoE on total CO₂eq (%)",
              ylim = c(0, max(df$moe) * 1.2), cex.names = 0.95,
              cex.axis = 0.85, cex.lab = 0.9)
text(bp, df$moe + max(df$moe) * 0.04, sprintf("%.1f%%", df$moe),
     cex = 0.85, font = 2)
legend("topleft", fill = c("#94A3B8", "#2D6A4F"),
       legend = c("No correlations", "Structural defaults"),
       border = NA, bty = "n", cex = 0.85)
```

## The seven pairs in the structural-defaults preset

When you pick "Structural defaults" on the left-hand card, seven specific parameter pairs get correlated. Each one is anchored in published livestock science:

- **BW ↔ MW** at ρ = +0.50. A heavier-than-average adult cow was probably a heavier-than-average juvenile too. From Brody's growth-curve work in the 1940s and the IPCC Eq 10.3 maintenance-energy formula.

- **BW ↔ WG** at ρ = +0.40. Well-fed adult cattle that are both heavier and gaining more — typical of finishing systems. From NRC 2001 joint regression estimation of body weight and weight gain. The sign is system-dependent; for growing-young-stock systems it can flip negative.

- **Milk ↔ Fat** at ρ = −0.30. The milk-dilution effect — when a cow makes more litres, the fat content per litre drops. From Wilmink (1987), VandeHaar (1998), and the standard dairy genetics literature.

- **Milk ↔ BW** at ρ = +0.30. High-producing dairy breeds are physically larger. A Holstein (650 kg, 30 kg milk/day) versus a Jersey (450 kg, 20 kg milk/day) is the textbook contrast. NRC Dairy 2001.

- **Milk ↔ DE** at ρ = +0.30. Higher feed digestibility lifts nutrient availability, which lifts milk yield. Strongest in concentrate-fed dairy. NRC Dairy 2001 Ch. 2.

- **DE ↔ CP** at ρ = +0.50. Feed-quality co-variation. High-quality forages tend to have both high digestibility and high crude protein. From NRC, INRA, and Feedipedia cross-feedstuff data.

- **DE ↔ Y~m~** at ρ = −0.50. The big one. IPCC 2019 Refinement Eq 10.21 makes Y~m~ an explicit decreasing function of `DE`. Observational meta-analyses (Niu et al. 2018, Hristov et al. 2013) confirm the slope. This is the pair that moves the headline the most in dairy systems.

Here's the matrix laid out visually:

```{r preset-heatmap, fig.width = 8, fig.height = 6, fig.alt = "Heatmap of the seven structural-defaults correlation pairs"}
m <- runs$country_x_preset$ad_corr_matrix %||% runs$country_x_preset$corr_matrix
if (!is.null(m)) {
  off_diag <- m; diag(off_diag) <- 0
  has_corr <- rowSums(abs(off_diag)) > 0 | colSums(abs(off_diag)) > 0
  if (sum(has_corr) >= 2) m <- m[has_corr, has_corr]
  par(mar = c(5, 5, 1, 1))
  image(seq_len(ncol(m)), seq_len(nrow(m)), t(m[nrow(m):1, ]),
        zlim = c(-1, 1), axes = FALSE, xlab = "", ylab = "",
        col = colorRampPalette(c("#C0392B", "white", "#2D6A4F"))(99))
  axis(1, at = seq_len(ncol(m)), labels = colnames(m), las = 2, cex.axis = 0.9)
  axis(2, at = seq_len(nrow(m)), labels = rownames(m)[nrow(m):1], las = 1,
       cex.axis = 0.9)
  for (i in seq_len(nrow(m))) for (j in seq_len(ncol(m))) {
    v <- m[nrow(m) + 1 - i, j]
    if (abs(v) > 0.01)
      text(j, i, sprintf("%.2f", v),
            col = if (abs(v) > 0.5) "white" else "black", cex = 0.85)
  }
}
```

Green = positive (move together), red = negative (move opposite), white = independent. Notice how sparse it is. Most off-diagonal cells are white because the tool deliberately keeps the preset minimal — only the seven pairs with published evidence are non-zero. Everything else is assumed independent until you tell the tool otherwise.

## Which of the four AD modes should you pick

The left-hand card on the Simulate tab has four options. In one paragraph each, here's how to choose:

**No correlations.** The strict IPCC Approach-2 default. Always defensible. Pick it when you want the conservative baseline, or when your activity data sources are all genuinely independent.

**Structural defaults (expert-elicited).** Recommended for single-year inventories. The seven biology pairs above. Pick it when you're doing a one-year national inventory and you don't have a multi-year time series of feed quality / herd structure. Most users land here.

**From template (auto, time-series).** Pick this if you have at least five years of populated `Parameter_TimeSeries` data in your upload. The tool computes Spearman rank correlations directly from your country's history (with first-difference detrending applied internally to strip out shared long-run growth). This is the most data-driven option, but it needs the data to actually be there — the radio greys out if your sheet is empty.

**Manual (CSV upload).** Pick this only if you have a published correlation matrix from a study similar to your inventory, or a result from an expert-elicitation exercise. The tool ships an example template so you don't have to type the matrix from scratch.

## The other card: emission-factor correlations

The Simulate tab actually has **two** correlation toggles, side by side. Everything above is about the **left-hand card** — correlations between **activity data** (herd size, body weight, milk yield, feed quality — quantities measured from the population). The **right-hand card** is for something different: correlations between **emission-factor coefficients** (Y~m~, B~o~, MCF, EF3, EF4, EF5 — numbers that come from published studies, not from your survey).

Why are they separate? Because the physical processes are different. Activity data co-varies because animals and feed move together in nature. Emission factors co-vary if and only if their published values share a common methodological bias — for example, when all your rumen-fermentation coefficients came from a single regional database, so they're all calibrated against the same dataset. That's a different kind of correlation, and it gets its own card.

### The three blocks

The right-hand card splits the IPCC coefficients into three groups based on which scientific literature they come from. Within a group, you can dial up a shared bias. Across groups, the tool always assumes independence (the three literatures don't talk to each other).

| Block | Coefficients | What they have in common |
|---|---|---|
| **Energy** | Cfi, Ca, C, Cp, Y~m~ | All from rumen-fermentation and animal-energetics studies. IPCC Vol.4 Ch.10 §10.2. If you got them all from one regional rumen database, they share that database's calibration choices. |
| **Manure-CH₄** | Bo, MCF, ASH | All from biogas-potential and lagoon studies. If you got them all from one anaerobic-digestion lab, they share its conditions. |
| **Manure-N** | EF3, EF4, EF5, Frac_GasMS, Frac_LeachMS, UE | All from soil-emissions and N-volatilisation studies. If you got them all from one soil-emissions modelling group, they share its calibration. |

### What the ρ slider does

When you pick "Block-structured EF correlation" and move the energy-block slider to ρ = 0.3, the sampler treats all five energy coefficients as if they share a common bias term with 30% strength. If Y~m~ samples high in a given iteration, Cfi gets nudged up too — but only weakly. Here's what that looks like for the Y~m~ / Cfi pair specifically:

```{r fig-ef-scatter, fig.width = 9, fig.height = 3.2, fig.alt = "Three scatter plots showing Ym vs Cfi under EF block correlation of 0, 0.3, and 0.5"}
set.seed(89)
n <- 1200
ym_marg  <- function(n) mc2d::rpert(n, min = 5.5, mode = 6.5, max = 7.5, shape = 4)
cfi_marg <- function(n) rnorm(n, mean = 0.386, sd = 0.024)
par(mfrow = c(1, 3), mar = c(3.5, 3.5, 2.5, 0.6), mgp = c(2.3, 0.7, 0))
for (rho in c(0, 0.3, 0.5)) {
  ic <- .ic_pair(n, rho = rho, marg1 = ym_marg, marg2 = cfi_marg)
  ttl <- if (rho == 0) "ρ = 0   (independent)" else sprintf("ρ = +%.1f", rho)
  plot(ic, pch = 16, col = adjustcolor("#2D6A4F", 0.4), cex = 0.5,
       xlab = "Yₘ (% gross energy)", ylab = "Cfi (MJ/kg^0.75/day)",
       main = ttl, cex.main = 0.95, cex.lab = 0.9, cex.axis = 0.85,
       xlim = c(5.3, 7.7), ylim = c(0.32, 0.45))
}
```

At ρ = 0, the cloud is round — Y~m~ and Cfi sampled independently. At ρ = 0.3, you can see a faint upward tilt. At ρ = 0.5, the tilt is obvious. The same logic applies to the other coefficients in the energy block; the slider lifts all of them together by the same factor.

### What it does to the headline

This is the demonstration on Country X. Same dairy inventory, varying which correlations are switched on:

```{r fig-ef-stacking, fig.width = 8, fig.height = 4.0, fig.alt = "Bar chart showing Country X 95% MoE under combinations of AD correlation and EF correlation modes"}
get_moe <- function(key) {
  s <- runs[[key]]$summary
  s[s$variable == "total_co2e", "moe_pct"]
}
df <- data.frame(
  config = c(
    "Nothing on",
    "EF only, ρ=0.3",
    "EF only, ρ=0.5",
    "AD preset only",
    "AD preset + EF ρ=0.3"
  ),
  moe = c(
    get_moe("country_x_none"),
    get_moe("country_x_ef_energy_03"),
    get_moe("country_x_ef_energy_05"),
    get_moe("country_x_preset"),
    get_moe("country_x_preset_plus_ef_03")
  ),
  ad  = c("none", "none", "none", "preset", "preset"),
  ef  = c("0", "0.3", "0.5", "0", "0.3")
)
par(mar = c(5.5, 4, 1.5, 0.6), mgp = c(2.6, 0.7, 0))
cols <- c("#94A3B8", "#94A3B8", "#94A3B8", "#2D6A4F", "#2D6A4F")
bp <- barplot(df$moe, names.arg = df$config, col = cols, border = "white",
              ylab = "95% MoE on total CO₂eq (%)",
              ylim = c(0, max(df$moe) * 1.18),
              las = 2, cex.names = 0.78, cex.axis = 0.85, cex.lab = 0.9)
text(bp, df$moe + max(df$moe) * 0.04, sprintf("%.1f%%", df$moe),
     cex = 0.82, font = 2)
```

The energy-block EF correlation on its own (ρ = 0.3) lifts the MoE a couple of percentage points above the no-correlation baseline. At ρ = 0.5 it lifts a bit more. Stack it on top of the structural-defaults AD preset and the effects mostly add — but only a little, because the AD effect already moved you partway and the EF effect is smaller in magnitude.

### When you should — and shouldn't — engage it

**Leave it off** if your emission factors come from IPCC defaults. The IPCC defaults are aggregated across many studies; no single methodological bias dominates them.

**Leave it off** if your EFs come from country-specific measurements done at different labs in different years. The errors there are roughly independent.

**Turn the energy block to ρ = 0.3** if all your energy-equation coefficients (Cfi, Ca, Y~m~, etc.) come from a single national feeding trial or one regional rumen database. ρ = 0.3 reflects a moderate shared bias, the kind that should give you pause but not full alarm. The "Currently:" caption under each slider tells you what band you're in: 0 is independent, 0.15 is weak, up to 0.35 is moderate, above 0.35 is strong (and should be justified in writing if you publish).

**Same logic for the other two blocks.** Manure-CH~4~ block — engage if all your Bo / MCF / ASH numbers came from one anaerobic-digestion lab. Manure-N block — engage if all your N-volatilisation parameters came from one soil-emissions modelling exercise. Otherwise leave them at 0.

If you toggle "Block-structured" on but leave all three sliders at 0, an inline amber warning appears under the sliders telling you the choice has no effect — heed it.

## Common mistakes

Things that catch people out, and quick fixes:

- **Expecting the mean to change.** It doesn't. Correlations widen or narrow the spread; they preserve the central estimate to within Monte Carlo noise. If you change correlation modes and the mean shifts more than a few tenths of a percent, you have a different problem (probably parameter rebroadcasting, not correlation).

- **Picking "From template (auto)" on an empty `Parameter_TimeSeries` sheet.** Before mid-2026 this silently fell back to no correlations. The radio now greys out automatically when the sheet is empty, and the Run button blocks with a clear error if a stale state slips through. If you see grey-vs-green bars sitting at identical heights on the comparison chart, you're probably in this case.

- **Picking "Block-structured EF correlation" with all three sliders at 0.** Same silent-no-op trap on the right-hand card. The amber warning underneath the sliders is your cue — move at least one slider, or switch back to "No EF correlations".

- **Renaming row/column headers in the manual matrix.** Names that don't match the canonical `PARAM_CATALOGUE` are silently dropped. Always start from the downloadable example template — the canonical names are already in place.

- **Worrying about positive-definiteness.** If your uploaded matrix isn't positive-definite, the tool projects it onto the nearest positive-definite correlation matrix automatically and warns you if any entry shifts by more than 0.01. You don't have to construct a perfectly valid matrix by hand.

- **Reading a small effect as broken.** A single correlated pair inside a 13-input multiplicative chain only moves the headline 95% MoE by a few percent. That's the correct mathematical behaviour, not a bug. If you want a larger effect you need either richer multi-year time-series data (the "From template" mode harvests many correlated pairs at once) or genuinely wide per-parameter uncertainty bounds.

## See it on your own data

On Tab 5, tick the **"Compare with/without correlations"** checkbox before pressing Run. The Results tab will then show a green-vs-grey bar chart titled *"Effect of Correlations on Uncertainty"*. Green is your selected correlation mode; grey is the same simulation with all correlations turned off. The gap between them is the actual effect of correlations on *your* inventory.

That's the most honest test. The Country X / Country Y examples above show how it can play out, but your inventory's per-parameter bounds and sub-category mix will produce its own characteristic gap.

<div class="related">
### Related material

- [Methodology document](../methodology.pdf), §Correlation Structure — the full equations behind each preset pair, the Iman–Conover sampling procedure, and the rejected pairs (N↔BW, Cfi↔Ca, Milk↔pct_pregnant) with reasons.
- [User Guide](../user_guide.pdf), §Correlations (Simulate tab) — step-by-step workflow including how to download the manual-entry CSV template.
- IPCC 2006 Vol.1 Ch.3 §3.2.3.2 — the restricted-pairing rank-correlation procedure (Iman & Conover 1982) that the tool implements.

</div>

<footer>
Generated by <code>scripts/build_help_docs.R</code> from <code>docs/correlations.Rmd</code>. Worked-example numbers reflect the Country X / Country Y built-in datasets at the current app version.
</footer>
